Assessing the Population-level Conservation Effects of Marine Protected Areas

### 2020-07-17 documents/ovando-et-al-pop-effects-of-mpas-si.Rmd
starting final word sladh, need to lose 1600 :(

---
title: "Supporting Information"
subtitle: "Assessing the Population-level Conservation Effects of Marine Protected Areas"
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
output: 
  bookdown::pdf_document2:
    keep_tex: true
  pdf_document: default
bibliography: My Library.bib
header-includes:
    - \usepackage{longtable}
    - \usepackage{booktabs}
    - \usepackage{eso-pic}
    - \usepackage{graphicx}
    - \usepackage[left]{lineno}
    - \usepackage{xcolor}
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
always_allow_html: true
---

@@

```{r, include=FALSE}
library(sf)
library(ggmap)
library(viridis)
library(hrbrthemes)
library(scales)
library(patchwork)
library(rpart)
library(extrafont)
library(caret)
library(ggsci)
library(rEDM)
library(tidyverse)
library(ggtext)
library(rstan)
library(rstanarm)
library(bayesplot)
library(Synth)

@@

@@

# SI Text {#si-text .unnumbered}

@@

\newpage
```{r cartoon, fig.cap= "Cartoon illustration of the hierarchichal difference-in-difference estimator", fig.width=3}
knitr::include_graphics("figs/estimation-cartoon.png")
```

>>>>

---
title: "Online Supporting Information"
subtitle: "Assessing the Population-level Conservation Effects of Marine Protected Areas"
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
output: 
  bookdown::pdf_document2:
    keep_tex: true
  pdf_document: default
bibliography: references.bib
header-includes:
    - \usepackage{longtable}
    - \usepackage{booktabs}
    - \usepackage{eso-pic}
    - \usepackage{graphicx}
    - \usepackage[left]{lineno}
    - \usepackage{xcolor}
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
always_allow_html: true
---

@@

```{r, include=FALSE}
library(sf)
library(ggmap)
library(viridis)
library(hrbrthemes)
library(scales)
library(patchwork)
library(rpart)
library(extrafont)
library(caret)
library(ggsci)
library(rEDM)
library(tidyverse)
library(ggtext)
library(rstan)
library(rstanarm)
library(bayesplot)
library(Synth)
library(here)

@@

# (APPENDIX) Appendix {-}

@@

@@

\newpage
```{r cartoon, fig.cap= "Cartoon illustration of the hierarchichal difference-in-difference estimator", fig.width=8}
knitr::include_graphics(here::here("documents","figs","estimation-cartoon.png"))
```

### 2020-07-17 documents/ovando-et-al-pop-effects-of-mpas.Rmd
starting final word sladh, need to lose 1600 :(

---
title: |
  | Assessing the Population-level Conservation Effects of
  | Marine Protected Areas
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world’s ocean, up from less than 1% in the year 2000, and international commitments call for 10%, 30%, and even 50% coverage. The premise underlying MPA expansion is that they conserve biodiversity, habitats, and fished populations. While numerous studies show that MPAs produce conservation benefits *inside* their borders, many MPAs are also justified on the grounds that they confer conservation benefits to the broader population *beyond* their borders. This paper examines the conditions under which MPAs can provide population-level conservation benefits inside and outside their borders, and shows that even in cases where the population benefits are large, they are inherently difficult to detect empirically. A network of MPAs was put in place in The Channel Islands National Marine Sanctuary in 2003, with a goal of providing regional conservation and fishery benefits. Evidence indicates that the Channel Island MPAs have increased biomass densities inside the MPAs, but we are unable to find a clear effect of these same MPAs at the population level. MPA effect sizes less than 30% are likely to be difficult to detect (even when they are present), and the size of many MPA networks suggests that effect sizes may often be smaller than 30%. Our results provide a novel assessment of the population-level effects of a large and iconic Marine Protected Area network, and provide guidance for communities charged with monitoring and adapting MPAs.
bibliography: My Library.bib
csl: conservation-biology.csl
output:
  bookdown::word_document2:
    reference_docx: template.docx
  bookdown::pdf_document2: default
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
always_allow_html: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 600,
                      cache = FALSE, fig.width = 3.42, fig.asp = 1.5, dev = "png")

@@

# Front Matter

## Author Contributions: 
  D.O., S.G., and R.H. developed structure of simulation model. D.O., O.D., C.C., and J.C. developed estimation strategy. J.C. provided support on collecting and interpreting data. All simulations, statistics, and sensitivity analyses performed by D.O. and O.L. . All authors contributed to the manuscript.

## Conflicts of Interest:
  R.H's research program receives funding from environmental NGOs, foundations, fishing industry, governments and international agencies. All of these can be interpreted as a conflict of interest when evaluating fisheries policy. C.C., S.G., and O.D's research program includes funding from environmental NGOs and foundations with an interest in ocean conservation and food security. 
## Significance Statement
  Healthy marine ecosystems are critical to the well-being of the planet. Marine protected areas (MPAs) are increasingly being used to conserve and manage these ecosystems. We pair theory with empirical methods to show that the population-level conservation effects of marine protected areas can be highly variable and challenging to detect. We find no statistically clear effect of a large and well protected network of MPAs in the Channel Island, California, on targeted fin-fish after 14 years of protection, which our simulation analysis suggests is in fact to be expected. MPA practitioners must clearly communicate expectations from MPA networks, and plan monitoring and adaptations strategies that directly address the challenges of estimating population-level conservation effects of protected areas.

## Acknowledgements
  Funding for the simulation model was provided by the NMFS-Sea Grant Population and Ecosystem Dynamics Fellowship. Empirical data collection was funded primarily by the David and Lucille Packard Foundation in support of the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) with additional funding from the California Ocean Protection Council and California SeaGrant. This study would not be possible without the work provided by PISCO divers over the years, with special thanks to to Kathyrn Davis Koehn and Avrey Parsons-Field. We also thank Cody Szuwalski, Julia Lawson, André Punt, and two anonymous reviewers for helpful comments and technical support.

## Keywords:
  - Marine Protected Area Networks
  - Conservation
  - Bio-economic modeling
  - Program Evaluation

@@

The empirical MPA literature has generally focused on assessing conservation effects within MPA borders [@lester2009;@halpern2003; @edgar2014; @gerber2005].As conservation benefits accrue inside MPAs, MPAs can affect the waters beyond their borders through the spillover of adult and larval fish from the protected to the fished areas, as well as through displacement of fishing effort. These dynamics link the state of fishes inside MPAs to those outside. Several studies have documented empirical evidence for the existence of adult or larval fish spillover affecting both fish abundance and fisheries productivity [e.g. @goni2010; @halpern2009;@kay2012; @stobart2009; @mcclanahan2000; @russ1996; @thompson2017;@baetscher2019;@pelc2009]. MPAs can also cause alterations in fishing effort [@costa2013;@mason2012;@murawski2005;@mcdermott2019].

Numerous factors influence how MPAs affect region-wide fish populations. These include the scale of adult and larval dispersal relative to the size of the MPAs  [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008;@gerber2005], the strength and timing of density dependence [@burgess2014], the degree of enforcement [@edgar2014], how overfished the population would be without the MPA, the time span under evaluation [@ovando2016;@moffitt2013], and how fishing and management responds to the implementation of the MPAs [@hilborn2004; @hilborn1992; @walters2004; @hastings2003;@gerber2003; @hilborn2004a; @gaines2010; @botsford2003;@smith2003a;@smith2006a;@walters2000a;@white2011,;@white2010c]. In addition, even for the same total area of MPAs, the location and spacing of the MPAs can have a profound influence on their cumulative impact through network effects [@gaines2010; @costello2010a].

The empirical evidence of the conservation outcomes of MPAs is made up almost entirely by versions of "response ratios", commonly measured as the ratio of biomass densities of species inside relative to outside of MPAs [@lester2009; @halpern2003a;@edgar2014;@caselle2015]. These studies have by and large found clear evidence that well enforced and sufficiently sized MPAs are associated with high response ratios. Why does this large body of empirical work not serve as sufficient evidence for the population-level effects of MPAs?

Control sites used in calculating response ratios are often selected based on habitat characteristics [@ferraro2018]. However, export of adults or larvae from the MPA to these "control" site affects their status as controls, as does displacement of fishing effort from MPAs to control sites. In theory, control sites far enough away to negate both biological spillover and concentration by the fishing fleet could be selected, but finding suitably far sites that are also appropriate proxies for the ecological and economic context of the MPAs is challenging. As a result, response ratios can be a highly imprecise and biased measure of the population-level conservation effect of an MPA network [@ferraro2018]. This means that our collective scientific understanding of the regional-scale impacts of MPAs is surprisingly limited.

@@

The MPAs within the Channel Islands National Marine Sanctuary, California, USA (hereafter the Channel Islands) provide an opportunity to improve our understanding of population-level MPA effects. A network of protected areas covering approximately 20% of the Islands' waters was put in place in 2003 [see @osmond2010, @kirlin2013, @botsford2014, and @hilborn2012 for information on the creation of these MPAs]. The network has been used as a model case study in protected area design around the world [@botsford2014 and references therein]. We use data from the first 14 years of protection to provide what is to our knowledge the first empirical assessment of the population-level effect of a large MPA network on a wide array of fin-fish species. In contrast to clear differences in biomass densities observed inside and outside of well-protected MPAs both globally [@lester2009] and in the Channel Islands [@caselle2015] we are unable to detect a clear population effect from the Channel Islands MPAs. We build off of existing MPA theory to consider why this might be, and provide guidance for scientists and managers as to when and how we might expect to estimate the population-level conservation effects of MPAs.

@@

Our primary methods consist of a spatially-explicit bio-economic simulation model and a Bayesian difference-in-difference regression. We use the simulation model to provide provide theoretical grounding and context for our empirical results. The difference-in-difference regression provides our estimate of the population-level conservation effect of the Channel Islands MPAs. All analysis were conducted in R [@rcoreteam2019]. Our difference-in-difference model was fit using Stan [@carpenter2017] using the `rstanarm` package [@goodrich2020]. All data and code needed to fully reproduce our results are publicly available at [github.com/DanOvando/population-effects-of-mpas](https://github.com/DanOvando/population-effects-of-mpas).

@@

Our bio-economic model simulates the effect of MPAs on a spatially explicit age-structured representation of a fish population. Readers can explore the functionality of the model using an online tool available at [danovando.shinyapps.io/simmpa/](https://danovando.shinyapps.io/simmpa/). Detailed methods can be found in the online appendix.

@@

where $d_i$ is the biomass density at observation *i*, *T* indicates whether the observation *i* is for a targeted ($T = 1$) or non-targeted ($T = 0$) species, and *MPA* marks whether observation *i* is in a pre MPA ($MPA = 0$) or post MPA ($MPA = 1$) state. **B^c^** is a vector of coefficients for additional control variables in matrix *X* such as water visibility and observer experience. **B^s^** is a vector of hierarchical coefficients for each sampling location **S** , clustered by island.  Under the assumptions of this model, $\beta_3$ is the causal effect of the treatment (*MPA*) on the treated (targeted species). $\beta_0$ is the mean total biomass of non-targeted species pre-MPA, and $\beta_1T$ is the mean total biomass of targeted species pre-MPA. $\beta_2MPA$ is the effect of the post-MPA period on the non-targeted species. The shape and scale parameters of the Gamma distribution are estimated as well. Regression details and diagnostic plots are available in the online appendix. The full table of covariates and their posterior means and 89% credible interval can be found in Table.\@ref(tab:coef-table).

@@

We compared this range of plausible outcomes to the empirical data. Previous research [@caselle2015] found a statistically significant increase in the response ratios of targeted species over time, and evidence that this increase is smaller in the non-targeted species. Updating the results of @caselle2015 through 2017 shows a continuation of the increasing trend in the response ratios of targeted species (Fig.\@ref(fig:rr-plot)). This provides evidence that the Channel Islands MPAs are large enough and sufficiently well-enforced as to provide meaningful protection within their borders.

@@

While we do not have access to fine scale fishing data from the Channel Islands alone, reported catches for the species of interest in the Santa Barbara region exhibit a mix of stable, downward, and upward trajectories (see SI Appendix Fig.S1) which indicates that a negative MPA effect caused by a constant-catch fishing strategy is unlikely. However, some of the most common targeted species in the Channel Islands, such as blue rockfish and copper rockfish, have seen dramatic increases in reported catches in the Santa Barbara region since the year 2010 (although other such as California sheephead have seen declines relative to pre-MPA catch level). To the extent that these increases in catches are due to external market forces, the disappearance of clear positive population-level MPA effects estimated by our model could be due to exogenous increases in fishing pressure overwhelming any population-level increases caused by the MPA network. Unfortunately, we lack a control that would allow us to estimate the MPA effects relative to a counterfactual in which the same market forces increase catches but no MPAs are in place.

@@

Trophic cascades are another possible explanation for our equivocal results. All of the species in this empirical analysis may affect each other through mechanisms such as predation, competition, and habitat modification. We used convergent cross mapping (CCM), in the manner of @clark2015, to test for significant dynamic interactions between species and therefore the possibility of the trophic cascades biasing our results. Our results found no significant cross-mappings between targeted and non-targeted species, indicating that while clearly there are interactions between these groups on some level, the effects within the timespan of the data are not pronounced enough to be of concern to our results (SI Appendix Fig.S42). However, the longer MPAs are in place, the greater the possibility that substantial species interactions that can affect use of non-targeted species as a control may arise.

@@

<!-- Once simulation models have been used to help design an MPA, how can users evaluate whether it is achieving their objectives?  -->
How then should stakeholders reconsider the population-level effects of MPAs? Response ratios are commonly used as evidence for conservation outcomes of MPAs; @lester2009 and @halpern2002 present meta-analyses of hundreds of such studies. These results often find massively higher densities and biomass inside MPAs than outside [median biomass increases of  over 400%, @lester2009]. But as suggested in @white2011, @ferraro2018, @larsen2019, @moffitt2013 and further demonstrated here, when MPAs affect nearby control sites used in response ratios through biological spillover or concentration of fishing effort, it is entirely possible to for MPA to produce massive response ratios while simultaneously having minimal effects on the entire population partially protected by the MPAs (and *vice versa*).

As @larsen2019 suggests, there are many potential alternatives for estimating the effects of MPAs that better account for the challenges of causal inference (though that may be more data-intensive). We applied one such approach here (a difference-in-difference estimator, see online appendix for alternative strategies such as synthetic controls [@abadie2010] that show similar results), and yet were still unable to reach robust conclusions as to the effect of MPAs on the total biomass density of targeted fin-fish in the Channel Islands, due to the likely small size of the true effect relative to the influence of environmental variability.

@@

Non-equilibrium analyses also help set expectations for effect sizes over time [@nickols2019,@kaplan2019,@ovando2016,Fig.\ref{did-plot}]. Educating communities about the challenges of estimating the effects of MPAs can help ensure that a lack of a clear effect is not necessarily viewed as a failure of the program, or large positive result based solely on response ratios as a clear sign of success and subsequent relaxation of other fishery management strategies. Rather, results and subsequent management actions must be considered in the context of reasonable expectations given the size, age, and degree of enforcement of the MPAs in question, together with the ecological and economic dynamics of a given system. While this paper has focused on the conservation outcomes of MPAs, future work must also address the challenge of predicting and estimating the fishery impacts of protected areas [in the manner of @smith2006a].

@@

# Supporting Information

A detailed description of the simulation model structure, as well as sensitivity analyses of our estimation model, are available in the online appendix.

@@

knitr::kable(equations,
             escape = FALSE,
             caption = "Variables for MPA simulations. U mean uniform",
             format = "pipe") %>% 
  kableExtra::kable_styling()

@@

```{r rr-plot, include = TRUE, fig.width=width_1col, fig.asp=asp_1col,,fig.cap="90% Posterior probability distributions of response ratios for targeted species (x-axis) over time (y-axis) in blue. Simulated population-level (pop.) effect on biomass densities matched to empirical response ratios in red. For response ratios, a value of zero indicates that biomass densities of targeted species are identical inside and outside MPAs, a value of one that biomass densities of targeted species are 100% inside MPAs relative to outside. For MPA population effect, a value of zero indicates that biomass densities are identical in the with- and without- MPA scenarios. A value of 1 indicates that biomass densities are 100% greater in the scenario with MPAs than the scenario without MPAs."}

@@

# References

>>>>

---
title: Assessing the Population-level Conservation Effects of Marine Protected Areas
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world’s ocean, up from less than 1% in the year 2000, and international commitments call for 10%, 30%, and even 50% coverage. The premise underlying MPA expansion is that they conserve biodiversity, habitats, and fished populations. While numerous studies show that MPAs produce conservation benefits *inside* their borders, many MPAs are also justified on the grounds that they confer conservation benefits to the broader population *beyond* their borders. This paper examines the conditions under which MPAs can provide population-level conservation benefits inside and outside their borders, and shows that even in cases where the population benefits are large, they are inherently difficult to detect empirically. A network of MPAs was put in place in The Channel Islands National Marine Sanctuary in 2003, with a goal of providing regional conservation and fishery benefits. Evidence indicates that the Channel Island MPAs have increased biomass densities inside the MPAs, but we are unable to find a clear effect of these same MPAs at the population level. We show that MPA effect sizes less than 30% are likely to be difficult to detect (even when they are present); the size of many MPA networks suggests that effect sizes may often be smaller than 30%. Our results provide a novel assessment of the population-level effects of a large and iconic Marine Protected Area network, and provide guidance for communities charged with monitoring and adapting MPAs.
bibliography: references.bib
csl: conservation-biology.csl
output:
  bookdown::word_document2:
    reference_docx: template.docx
  bookdown::pdf_document2:
    keep_tex: true
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
always_allow_html: true
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, dpi = 300,
                      cache = FALSE, fig.width = 3.42, fig.asp = 1.5, dev = "pdf")

@@

@@

The empirical MPA literature has generally focused on assessing the ability of MPAs to provide conservation gains within their borders [@lester2009;@halpern2003; @edgar2014; @gerber2005]. However, as conservation benefits accrue inside MPAs, MPAs can affect the waters beyond their borders through the spillover of adult and larval fish from the protected to the fished areas, as well as through displacement of fishing effort. These dynamics result in MPAs affecting not only the organisms within their borders, but the broader population, the connected biological stock, of affected species. Theory states that numerous factors influence how MPAs might affect region-wide fish populations. These include the scale of adult and larval dispersal relative to the size of the MPAs [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008;@gerber2005], the strength, timing, and location of density dependence [@burgess2014], the degree of enforcement [@edgar2014], how overfished the population would be without the MPA, the time span under evaluation, and how fishing and management responds to the implementation of the MPAs [@hilborn2004; @hilborn1992; @walters2004; @hastings2003;@gerber2003; @hilborn2004a; @gaines2010; @botsford2003;@smith2003a;@smith2006a;@walters2000a;@white2011,;@white2010c;@ovando2016;@moffitt2013]. Even for the same total area of MPAs, the location and spacing of the MPAs can have a profound influence on their cumulative impact through network effects [@gaines2010; @costello2010a].

What empirical evidence is there for these population-level effects of MPAs? Conservation outcomes of MPAs are generally evaluated by versions of "response ratios", commonly measured as the ratio of biomass densities of species inside relative to outside of MPAs [@lester2009; @halpern2003a;@edgar2014;@caselle2015]. These studies have by and large found clear evidence that well enforced and sufficiently sized MPAs are associated with high response ratios. Several studies have also documented empirical evidence for the existence of adult or larval fish spillover affecting both fish abundance and fisheries productivity [e.g. @goni2010; @halpern2009;@kay2012; @stobart2009; @mcclanahan2000; @russ1996; @thompson2017;@baetscher2019;@pelc2009]. MPAs have also caused alterations in fishing effort [@costa2013;@mason2012;@murawski2005;@mcdermott2019]. Why does this large body of empirical work not serve as sufficient evidence for the population-level effects of MPAs?

While spillover-based studies provide evidence that MPAs may have effects outside their borders, they are not sufficient evidence for the net effect of the MPAs. Detection of a biomass density or catch-per-unit-effort gradient decreasing with distance from an MPA border suggests that there may be spillover, but does not tell us the net effect of that spillover on the population or the fishery. Response ratios have shortcomings as measures of population-level effects as well. Control sites used in calculating response ratios are often selected based on habitat characteristics [@ferraro2018]. However, export of adults or larvae from the MPA to these "control" site affects their status as controls, as does displacement of fishing effort from MPAs to control sites. In theory, control sites far enough away to negate both biological spillover and concentration by the fishing fleet could be selected, but finding suitably far sites that are also appropriate proxies for the ecological and economic context of the MPAs is challenging. As a result, response ratios can be a highly imprecise and biased measure of the population-level conservation effect of an MPA network [@ferraro2018;@moffitt2013]. As response ratios are our primary source of evidence for the conservation effects of MPAs, this means that our collective scientific understanding of the population-level impacts of MPAs is surprisingly limited.

@@

The MPAs within the Channel Islands National Marine Sanctuary, California, USA (hereafter the Channel Islands) provide an opportunity to improve our understanding of population-level MPA effects. A network of protected areas covering approximately 20% of the Islands' waters was put in place in 2003 (see @osmond2010, @kirlin2013, @botsford2014, and @hilborn2012 for information on the creation of these MPAs). The network has been used as a model case study in protected area design around the world (@botsford2014 and references therein). We use data from the first 14 years of protection to provide what is to our knowledge the first empirical assessment of the population-level effect of a large MPA network on a wide array of fin-fish species. In contrast to clear differences in biomass densities observed inside and outside of well-protected MPAs both globally [@lester2009] and in the Channel Islands [@caselle2015] we are unable to detect a clear population effect from the Channel Islands MPAs. We build off of existing MPA theory to consider why this might be, and provide guidance for scientists and managers as to when and how we might expect to estimate the population-level conservation effects of MPAs.

@@

Our primary methods consist of a spatially-explicit bio-economic simulation model and a Bayesian difference-in-difference regression. We use the simulation model to provide provide theoretical grounding and context for our empirical results. The difference-in-difference regression provides our estimate of the population-level conservation effect of the Channel Islands MPAs. All analysis were conducted in R [@rcoreteam2019]. Our difference-in-difference model was fit using Stan [@carpenter2017] using the `rstanarm` package [@goodrich2020]. All data and code needed to fully reproduce this manuscript are publicly available at [github.com/DanOvando/population-effects-of-mpas](https://github.com/DanOvando/population-effects-of-mpas). A detailed description of the simulation model structure, as well as sensitivity analyses of our estimation model, are available in the online supporting information (SI).

@@

Our bio-economic model simulates the effect of MPAs on a spatially explicit age-structured representation of a fish population. Readers can explore the functionality of the model using an online tool available at [danovando.shinyapps.io/simmpa/](https://danovando.shinyapps.io/simmpa/). Detailed methods can be found in the online supporting information.

@@

where $d_i$ is the biomass density at observation *i*, *T* indicates whether the observation *i* is for a targeted ($T = 1$) or non-targeted ($T = 0$) species, and *MPA* marks whether observation *i* is in a pre MPA ($MPA = 0$) or post MPA ($MPA = 1$) state. **B^c^** is a vector of coefficients for additional control variables in matrix *X* such as water visibility and observer experience. **B^s^** is a vector of hierarchical coefficients for each sampling location **S** , clustered by island.  Under the assumptions of this model, $\beta_3$ is the causal effect of the treatment (*MPA*) on the treated (targeted species). $\beta_0$ is the mean total biomass of non-targeted species pre-MPA, and $\beta_1T$ is the mean total biomass of targeted species pre-MPA. $\beta_2MPA$ is the effect of the post-MPA period on the non-targeted species. The shape and scale parameters of the Gamma distribution are estimated as well. Regression details and diagnostic plots are available in the SI. The full table of covariates and their posterior means and 89% credible interval can be found in Table.\@ref(tab:coef-table).

@@

We compared this range of plausible outcomes to the empirical data. @caselle2015 found a statistically significant increase in the response ratios of targeted species over time, and evidence that this increase is smaller for non-targeted species. Updating the results of @caselle2015 through 2017 shows a continuation of the increasing trend in the response ratios of targeted species (Fig.\@ref(fig:rr-plot)). This provides evidence that the Channel Islands MPAs are large enough and sufficiently well-enforced as to provide meaningful protection within their borders.

@@

While we do not have access to fine scale fishing data from the Channel Islands alone, reported catches for the species of interest in the Santa Barbara region exhibit a mix of stable, downward, and upward trajectories (see SI Fig.S1) which indicates that a negative MPA effect caused by a constant-catch fishing strategy is unlikely. However, some of the most common targeted species in the Channel Islands, such as blue rockfish and copper rockfish, have seen dramatic increases in reported catches in the Santa Barbara region since the year 2010 (although other such as California sheephead have seen declines relative to pre-MPA catch level). To the extent that these increases in catches are due to external market forces, the disappearance of clear positive population-level MPA effects estimated by our model could be due to exogenous increases in fishing pressure overwhelming any population-level increases caused by the MPA network. Unfortunately, we lack a control that would allow us to estimate the MPA effects relative to a counterfactual in which the same market forces increase catches but no MPAs are in place.

@@

Trophic cascades are another possible explanation for our equivocal results. All of the species in this empirical analysis may affect each other through mechanisms such as predation, competition, and habitat modification. We used convergent cross mapping (CCM), in the manner of @clark2015, to test for significant dynamic interactions between species and therefore the possibility of the trophic cascades biasing our results. We found no significant cross-mappings between targeted and non-targeted species, indicating that while clearly there are interactions between these groups on some level, the effects within the timespan of the data are not pronounced enough to be of concern to our results (SI Fig.S42). However, the longer MPAs are in place, the greater the possibility that substantial species interactions that can affect use of non-targeted species as a control may arise.

@@

<!-- Once simulation models have been used to help design an MPA, how can users evaluate whether it is achieving their objectives?  -->
How then should stakeholders evaluate the population-level effects of MPAs? Response ratios are commonly used as evidence for conservation outcomes of MPAs; @lester2009 and @halpern2002 present meta-analyses of hundreds of such studies. These results often find massively higher densities and biomass inside MPAs than outside [median biomass increases of  over 400%, @lester2009]. But as suggested in @white2011, @ferraro2018, @larsen2019, @moffitt2013 and further demonstrated here, when MPAs affect nearby control sites used in response ratios through biological spillover or concentration of fishing effort, it is entirely possible to for MPA to produce massive response ratios while simultaneously having minimal effects on the entire population partially protected by the MPAs (and *vice versa*).

As @larsen2019 suggests, there are many potential alternatives for estimating the effects of MPAs that better account for the challenges of causal inference (though that may be more data-intensive). We applied one such approach here (a difference-in-difference estimator, see SI for alternative strategies such as synthetic controls, @abadie2010, that show similar results), and yet were still unable to reach robust conclusions as to the effect of MPAs on the total biomass density of targeted fin-fish in the Channel Islands, due to the likely small size of the true effect relative to the influence of environmental variability.

@@

Non-equilibrium analyses also help set expectations for effect sizes over time (@nickols2019;@kaplan2019;@ovando2016;Fig.\ref{fig:mpa-effect}). Educating communities about the challenges of estimating the effects of MPAs can help ensure that a lack of a clear effect is not necessarily viewed as a failure of the program, or large positive result based solely on response ratios as a clear sign of success and subsequent relaxation of other fishery management strategies. Rather, results and subsequent management actions must be considered in the context of reasonable expectations given the size, age, and degree of enforcement of the MPAs in question, together with the ecological and economic dynamics of a given system. While this paper has focused on the conservation outcomes of MPAs, future work must also address the challenge of predicting and estimating the fishery impacts of protected areas, an important question for which extremely little empirical evidence is available [@smith2006a].

@@

# Literature Cited

<div id="refs"></div>

@@

knitr::kable(equations,
             escape = FALSE,
             caption = "Variables for MPA simulations",
             format = "pipe") %>% 
  kableExtra::kable_styling()

@@

```{r rr-plot, include = TRUE, fig.width=width_1col, fig.asp=asp_1col,,fig.cap="90% Posterior probability distributions of response ratios for targeted species (x-axis) over time (y-axis) in blue. Simulated population-level (pop.) effect on biomass densities matched to empirical response ratios in red. For response ratios, a value of zero indicates that biomass densities of targeted species are identical inside and outside MPAs, a value of one that biomass densities of targeted species are 100% greater inside MPAs relative to outside. For MPA population effect, a value of zero indicates that biomass densities are identical in the with- and without- MPA scenarios. A value of 1 indicates that biomass densities are 100% greater in the scenario with MPAs than the scenario without MPAs."}

@@
